The bhsdtr package: a general purpose method of Bayesian inference for Signal Detection Theory models

### 2020-01-13 inst/preprint/paper.tex
po recenzjach 1, zmiana gabor.RData

\abstract{We describe a novel method of Bayesian inference for
  hierarchical or non-hierarchical equal-variance normal Signal
  Detection Theory models with one or more criteria. The method is
  implemented as an open-source R package that uses the
  state-of-the-art platform, Stan, for sampling from posterior
  distributions. Our method can accommodate binary responses as well
  as additional ratings and an arbitrary number of nested or crossed
  random grouping factors. The SDT parameters can be regressed on
  additional predictors within the same model via intermediate
  unconstrained parameters, and the model can be extended by using
  automatically generated human-readable Stan code as a template. In
  the paper we explain how our method improves on other similar
  available methods, we give an overview of the package, demonstrate
  its ease of use by providing a real-study data analysis
  walk-through, and show that the model successfully recovers known
  parameter values when fitted to simulated data.}

@@

\begin{document}
\maketitle
Many tasks used in psychology studies are essentially classification
tasks. For example, in a memory study participants may be required to
decide if a given test item is old or new, or, in a perceptual study,
an object may be either a letter or a digit. If a task requires
classification, it is always possible that conclusions based on
accuracy or percent correct are invalid because the ability to
discriminate between stimulus classes (i.e., sensitivity) is
confounded with bias, which is a tendency to classify stimuli as
belonging to a particular class \cite{GreenSwets66}. In principle, any
effect that manifests itself in differences in classification accuracy
may reflect differences in sensitivity, bias, or both. Signal
Detection Theory provides a simple and popular solution to this common
and important problem.

However, because the SDT model is non-linear, variability in its
parameters due to factors which are usually study-specific and of no
direct interest to the researcher such as participants or items has to
be accounted for, otherwise the estimates of SDT parameters are
biased. As we explain later in this paper, none of the available
methods of inference for hierarchical SDT models that we are aware of
addresses this problem correctly in it's generality. Hence, our main
goal was to create a correct implementation of the general
hierarchical linear regression structure defined on SDT parameters. We
argue that the \code{bhsdtr} package for R \cite{rstatistical}
provides exactly such an implementation and we have made it publicly
available at \code{https://github.com/boryspaulewicz/bhsdtr}, together
with the annotated source code that was used to perform all the
analyses and produce all the figures presented in this paper.

@@

According to SDT, each stimulus in a classification task gives rise,
by some unspecified cognitive process, to a unidimensional internal
evidence value $s$ sampled from a distribution that depends on the
stimulus class. For historical reasons the two stimulus classes are
often referred to as "noise" and "signal", and task performance is
described in terms of hits, correct rejections, omissions, and false
alarms, but this terminology is appropriate only when the model is
applied to tasks that require detection, which is far from always
being the case. In the most widely used version of the model, shown in
Fig.~\ref{basicsdt}, the two evidence distributions are normal with
the same variance, which is usually fixed at unity to make the model
identifiable. The distance $d'$ between the means of the evidence
distributions represents sensitivity. Because normal distributions are
unbounded, $s$ is always ambiguous, and so a criterion $c$ placed on
the evidence axis has to be used to reach a binary decision. A
participant is assumed to decide that a stimulus belongs to the first
class (e.g., an old item) if $s < c$, or that it belongs to the second
class (e.g., a new item) if $s \geq c$. The location of the decision
criterion represents bias.

@@

Perhaps the simplest way of using this model is to fit it to observed
response counts and use the estimated $d'$ values in place of the
percent correct scores; if the model is correct, the resulting
performance measure is not contaminated by bias. Obviously, the model
may not be correct, which is one of the reasons why we focus more on
the generalized version shown in Fig.~\ref{sdtr} below.

@@

Ratings are accommodated by introducing additional criteria and
modelling a combined response $y$, which represents both the binary
classification decision and rating as a single number. The value of
$y$ increases with the strength of evidence in favor of the second
stimulus class. For example, if confidence is rated on a four-point
scale, then $y = 1$ when a subject decides that a stimulus belongs to
the first class with certainty $4$, and $y = 5$ when a subject decides
that a stimulus belongs to the second class with certainty $1$. More
formally, if $K$ is the number of possible combined responses, then a
subject is assumed to give response $k$ if
$s \in (\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the decision
criteria, with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$ and
$+\infty$ respectively.

There is a good reason to collect the ratings and use the generalized
SDT model from Fig.~\ref{sdtr}, even when neither the ratings nor the
placement of criteria are relevant to the research problem. When
$K = 2$ (no ratings), the SDT model fits perfectly, regardless of
whether it is a reasonably good approximation to reality, because the
data and the model have the same dimensionality. This makes the
generalization to the $K > 2$ case particularly important, as it is
only when $K > 2$ that the formal assumptions of the model (e.g.,
equal or unequal variance) can be tested\footnote{In contrast to the
  formal assumptions, a psychological interpretation of the SDT
  parameters can be tested even when ratings are not available, e.g.,
  by means of selective influence \cite{sternberg2001separate}}, which
is often done by comparing observed and predicted ROC curves.

@@

If data have a hierarchical structure, but variability due to
subjects, items, or other sampled factors is not accounted for,
estimates of average (fixed) effects are not guaranteed to be unbiased
and conclusions are not guaranteed to generalize to the target
population. The not uncommon practice of analyzing data aggregated
over sampled factors represents an extreme case of ignoring
hierarchical data structure. The invalidity of this approach in the
context of SDT was clearly illustrated with the results of
simulational studies by \citeA{morey2008problematic}; however,
strictly speaking, it requires such demonstrations only in order to
illustrate a mathematical truth in a specific case. Because SDT is a
non-linear model, by definition, when estimates of its parameters are
based on data aggregated over sampled factors (e.g., $d'$ estimated
for hits and false alarms averaged over subjects), the expected values
of these estimates (e.g., what the calculated $d'$ actually estimates)
are not in general true population averages (e.g., true $d'$ in some
condition). Such estimates are biased and inference about a target
population based on them is simply not valid. To give a concrete
example, consider two unbiased participants, one with $d'_1 = 2$ and
one with $d'_2 = 4$. Their expected average accuracy is given by
$(\Phi(d'_1/2) + \Phi(d'_2/2)) / 2 = .91$, which corresponds to
$d' = 2.68$, whereas their true average $d'$ is $3$.

@@

Firstly, the standard errors or credible intervals associated with the
regression coefficients do not reflect the uncertainty in the SDT
parameter estimates because the latter are treated as mere data
points. The precision of parameter estimates often varies between
participants, items, or conditions, but when the estimates are treated
as data points, no use is made of this information. Secondly,
regressing parameters on numeric predictors makes their estimates
dependent on the common regression structure, and so also on each
other, which can improve the quality of the estimates, just as
assuming that random effects are samples from a common distribution
may improve their estimates.

The aforementioned problems can be dealt with by supplementing the SDT
model with the hierarchical linear regression structure. This can only
be achieved if the model is substantially reparametrized.

\section{Hierarchical Signal Detection Theory in a constrained
  parameter space}

Both $d'$ and $c$ have the virtue of being directly interpretable in
terms of sensitivity and bias. However, both $d'$ and $c$ are
constrained: $d'$ is non-negative and, when there is more than one
criterion, the elements of the $\bm{c}$ vector are order restricted
($\bm{c}_{i+1} > \bm{c}_i$). Hierarchical linear regression structure
can only be defined on unconstrained parameters, because random
effects are assumed to be normally distributed and normal distribution
is unbounded and because fixed effects are represented by
unconstrained parameters. We provide examples of the problems that may
arise when the SDT model is not correctly reparametrized in the
following summary of two hierarchical SDT implementations.

We are aware of two failed attempts at creating a general-purpose
method of inference for hierarchical SDT models with ratings. One is
the Gibbs sampler proposed by \citeA{morey2008problematic} and the
other is the Hierarchical Meta-d' model (HMeta-d') proposed by
\citeA{hmetad}. The HMeta-d' model is a hierarchical version of the
meta-d' model \cite{maniscalco2012signal}, which in turn is a
generalization of the SDT model that allows for a separate
"meta-sensitivity" to account for possible discrepancies between a
binary stimulus classification (referred to as a type 1 task) and the
associated rating task (referred to as a type 2 or meta-cognitive
task). We consider HMeta-d' here because it reduces to the SDT model
with ratings when the type 1 and type 2 sensitivities are equal.

The Gibbs sampler created by \citeA{morey2008problematic} allows for
at most two sampled factors to have independent normally distributed
random effects on the evidence distribution means. Unlike $d'$, each
evidence distribution mean considered in isolation is an unconstrained
parameter, but the mean of the second evidence distribution is by
definition greater than ($d' > 0$) or equal to ($d' = 0$) the mean of
the first. The authors explicitly admit that their algorithm does not
enforce this restriction because, as they claim in the paper, it would
greatly complicate analysis. At the same time the authors fail to
mention that an immediate consequence of allowing for negative $d'$
values is that the resulting posterior distribution loses its intended
interpretation since it can have a non-zero mass when $d' < 0$. The
outermost criteria are fixed at $0$ and $1$, and the ordering
restriction is enforced by assuming that the likelihood is $0$
whenever $\bm{c}_{i+1} \leq \bm{c}_i$. As the authors explain, because
a sampled factor can have independent random effects on the evidence
distribution means, it can have an effect on all the criteria:
shifting both means by the same amount in the same direction is
equivalent to keeping the sensitivity intact, while shifting the
criteria relative to the evidence distributions. However, the elements
of the criteria vector cannot be affected differently by the same
sampled factor, which is an unrealistic restriction. Participants
differ in how they place the criteria just as they differ in their
sensitivity and it is quite impossible for all the participants to
place the criteria in the same distance from each other.

In HMeta-d' the hierarchical structure is restricted to normally
distributed random intercepts of one sampled factor. In the HMeta-d'
model the $d'$ parameter is allowed to assume negative values also,
but the most problematic aspect of this implementation is again the
representation of the criteria. Each element of the criteria vector
has an associated independent normal distribution, which does allow
for criteria random effects, but does not enforce the necessary
ordering restriction. The elements of this vector are sorted to obtain
another criteria vector, and it is this sorted criteria vector that is
used to model the conditional combined response
distributions. Consequently, the model does contain parameters
representing the actual, order-restricted criteria, but, because
sorting is not injective, the space of the actual criteria is only
loosely related (i.e., not isomorphic) to the space of the
unrestricted criteria vectors that are associated with the
hierarchical structure.

@@

Note that the normal distribution centered at the midpoint is merely a
mapping device, not a third evidence distribution, and that, for the
reasons that will soon become clear, it is wider than the two evidence
distributions. The mapping expressed by Eq.~\ref{eq:1} is an
isomorphism between the $R^{K-1}$ space and the space of
order-restricted criteria vectors. It's inverse is given by
$\gamma_i =\ln{(\int_{c_{i-1}}^{c_i} f(s) \,ds /
  \int_{c_{K-1}}^{\infty} f(s) \, ds)}$, where $f$ is the standard
normal probability density function. The elements of the $\bm{\gamma}$
vector correspond to relative distances between pairs of adjacent
criteria because their exponents represent the relative magnitudes of
areas under the standard normal curve, delineated by the pairs of
adjacent criteria:
$e^{\bm{\gamma}_i} / e^{\bm{\gamma}_j} = (\Phi(\bm{c}_i) -
\Phi(\bm{c}_{i-1})) / (\Phi(\bm{c}_j) - \Phi(\bm{c}_{j-1}))$. When
$K=2$, only $\bm{\gamma}_1$ is free to vary, and its value directly
represents the direction and magnitude of bias: $\bm{\gamma}_1$ is $0$
when the criterion is placed at the midpoint between the evidence
distributions; the more negative (positive) $\bm{\gamma}_1$ is, the
more the criterion is shifted to the left (right) of the midpoint.

@@

The formal definition of a Bayesian model is not complete without
providing fixed values of all the parameters that define prior
distributions. Specifying the priors on sensitivity effects does not
pose any special difficulties. The sensitivity of an unbiased
classifier given percent correct (pc) is given by
$2 \Phi^{-1}(\text{pc})$. When $p(stim = 1) = 0.5$, the greater the
bias, the lower the accuracy, meaning that an unbiased sensitivity is
a lower bound on sensitivity given percent correct. Let us assume that
the majority of subjects are expected to achieve percent correct
within the $.51$ to $.99$ range, with negligible bias. Since
$\ln(2\Phi^{-1}(.51)) = -2.99$ and $\ln(2 \Phi^{-1}(.99)) = 1.54$, a
reasonable weakly informative prior on $\delta$ is normal with mean
$(1.54 - 2.99) / 2$ and standard deviation $(1.54 + 2.99) / 2$, which
is the default prior on delta effects in the \code{bhsdtr} package.

@@

The \code{bhsdtr} package implements the model in the Stan modelling
language because it uses a state-of-the-art adaptive Hamiltonian Monte
Carlo sampling algorithm which often handles high-dimensional
correlated posteriors better than a Gibbs sampler. Our package is
essentially a collection of documented functions: The
\code{aggregate\_responses} function aggregates data as much as
possible for efficiency, but without distorting the hierarchical
structure. The \code{make\_stan\_model} function creates a model
definition in the Stan language. The Stan code produced by the
\code{make\_stan\_model} function can be fitted as is or modified by
the user if needed, e.g., to change the prior distributions or to drop
the equal variance assumption. The \code{make\_stan\_data} function
creates regression model matrices and other data structures required
by the model created using the \code{make\_stan\_model}
function. Finally, the \code{plot\_sdt\_fit} function can be used to
visually assess the fit of the model by creating publication-ready ROC
curve plots or response distribution plots with posterior predictive
intervals calculated for the chosen $\alpha$ level.

@@

We will now explain how to perform some of the essential steps of a
typical data analysis process, which at the very least will usually
involve preparing the data, creating the model code, fitting the
model, assessing the fit, and possibly converting the unconstrained
$\delta$ and $\gamma$ parameters to $d'$ and $c$.

@@

A model is fitted using the \code{stan} function from the \code{rstan}
package. This function requires a special list of data structures used
by the model as well as a model specification expressed in the Stan
language. Every model has some fixed effects structure since, even
when there are no predictors, the model parameters can be expressed as
regressed on a vector of ones (i.e., an intercept). However, many
models also have a hierarchical structure and, if that is the case,
this hierarchical structure has to be specified when using the
\code{make\_stan\_model} function. This is done by providing a list of
lists of R model formulae. Each list of model formulae is composed of
at least three elements and specifies the correlated random effects of
one sampled factor. The \code{group} element specifies the sampled
factor; the \code{delta} and \code{gamma} elements specify which
effects are assumed to vary between the levels of this sampled
factor. When \code{make\_stan\_model} is used without any arguments, it
specifies a model without any random effects. Fixed effects model
matrices are specified by providing a list with at least two model
formulae, named \code{delta} and \code{gamma}, to the
\code{make\_stan\_data} function that is described later in this
paper. Non-default priors can be specified by adding optional elements
to the random and fixed effects specification lists, as described in
the \code{make\_stan\_data} function documentation.

In the study in question there was only one sampled factor, i.e., the
participants. Because duration was a within-participant variable, in
principle its effect could vary between the participants for all the
SDT parameters. However, a preliminary data analysis indicated that
the barely noticeable 32 ms difference in duration seemed to affect
only the sensitivity. Thus, it was assumed that $\delta$ may depend on
duration and order, but $\gamma$ may only be affected by
order. Because duration was a within-subjects variable, its effect on
$\delta$ was assumed to vary between the participants, but the only
random effect associated with $\gamma$ was the intercept:

@@

The \code{make\_stan\_data} function creates fixed and random effect
model matrices based on the respective model formulae using dummy
contrast coding. Note that the implicit intercept was suppressed for
the $\delta$ model matrix. In this way, $\delta$ was estimated for
every duration $\times$ order condition. The resulting separate
intercepts and slopes parametrization makes it easier to calculate
arbitrary contrasts on posterior samples. A more standard
parametrization was used for the $\gamma$ parameter because it was
initially assumed that the criteria depend only on order, and so there
was only one contrast of interest for every element of the $\gamma$
vector. On the other hand, in such cases nested parametrization (with
separate intercepts and slopes for every condition) may be more
convenient if a researcher is interested in the actual criteria, as we
will later explain when introducing the \code{gamma\_to\_crit}
function. This example also illustrates how the separation of the
$\delta$ and $\gamma$ regression structures makes it possible to test
a broad class of linear models representing the dependence of the SDT
parameters on additional variables.

@@

Note that since more than one sampled factor is allowed, the names of
all the hierarchical parameters are indexed (e.g., \code{delta\_sd\_1},
\code{delta\_random\_1}, \code{Corr\_delta\_1}). The name
\code{counts\_new} refers to posterior predictive samples that are
required by the \code{plot\_sdt\_fit} function. Names starting with
\code{Corr} refer to random effects correlation matrices, which are
computed from Cholesky decompositions.

@@

Four chains of 8,000 iterations each were run simultaneously; the
first half of the posterior samples, which served as a warm-up period
for tuning the parameters of the sampling algorithm, was
discarded. Part of the resulting Stan output is presented in
Table~\ref{fitsum} below.

@@

>>>>

\abstract{We describe a novel method of Bayesian inference for
  hierarchical or non-hierarchical equal-variance normal Signal
  Detection Theory models with one or more criteria. The method is
  implemented as an open-source R package that uses the
  state-of-the-art platform, Stan, for sampling from posterior
  distributions. Our method can accommodate binary responses as well
  as additional ratings and an arbitrary number of nested or crossed
  random grouping factors associated with sensitivity or criteria
  effects. The SDT parameters can be regressed on additional
  predictors within the same model via intermediate unconstrained
  parameters, and the model can be extended by using automatically
  generated human-readable Stan code as a template. In the paper we
  explain how our method improves on other similar available methods,
  we give an overview of the package, demonstrate its ease of use by
  providing a real-study data analysis walk-through, and show that the
  model successfully recovers known parameter values when fitted to
  simulated data.}

@@

\begin{document}
\maketitle Many tasks used in psychology studies are essentially
classification tasks. For example, in a memory study participants may
be required to decide if a given test item is old or new, or, in a
perceptual study, an object may be either a letter or a digit. If a
task requires classification, it is always possible that conclusions
based on accuracy or percent correct are invalid because the ability
to discriminate between stimulus classes (i.e., sensitivity) is
confounded with bias, which is a tendency to classify stimuli as
belonging to a particular class \cite{GreenSwets66}. In principle, any
effect that manifests itself in differences in classification accuracy
may reflect differences in sensitivity, bias, or both. Signal
Detection Theory provides a popular solution to this common problem.

However, because the SDT model is non-linear, variability in its
parameters due to factors which are usually study-specific and of no
direct interest to the researcher such as participants or items has to
be accounted for. When they are not accounted for, e.g., by
aggregating the data accros participants or stimuli, the estimates of
SDT parameters are biased. As we explain later in this paper, none of
the available methods of inference for hierarchical SDT models that we
are aware of addresses this problem correctly in it's full generality,
meaning that none of the available methods allows for fixed and random
effects in both the sensitivity and the criteria parameters while
restricting the parameters in accordance with SDT assumptions. Later
in the paper we explain why the \code{bhsdtr} package for R
\cite{rstatistical}, which we have made publicly available at
\code{https://github.com/boryspaulewicz/bhsdtr}, provides a correct
implementation of the general hierarchical linear regression structure
defined on SDT parameters. The package repository also contains the
annotated source code that was used to perform all the analyses and
produce all the figures presented in this paper.

@@

According to SDT, each stimulus in a classification task gives rise,
by some unspecified cognitive process, to a unidimensional internal
evidence value $s$ sampled from a distribution that depends on the
stimulus class. For historical reasons the two stimulus classes are
often referred to as "noise" and "signal", and task performance is
described in terms of hits, correct rejections, omissions, and false
alarms, but this terminology is appropriate only when the model is
applied to tasks that require detection, which is far from always
being the case. In the most widely used version of the model, shown in
Fig.~\ref{basicsdt}, the two evidence distributions are normal with
the same variance, often fixed at unity to make the model
identifiable. The distance $d'$ between the means of the evidence
distributions represents sensitivity. Because normal distributions are
unbounded, $s$ is always ambiguous, and so a criterion $c$ placed on
the evidence axis has to be used to reach a binary decision. A
participant is assumed to decide that a stimulus belongs to the first
class (e.g., an old item) if $s < c$, or that it belongs to the second
class (e.g., a new item) if $s \geq c$. The location of the decision
criterion may be interpreted in terms of bias.

@@

Perhaps the simplest way of using this model is to fit it to observed
response counts and use the estimated $d'$ values in place of the
percent correct scores; if the model is correct, the resulting
performance measure is not contaminated by bias. The model may not be
correct, which in our view is the most important reason to focus more
on the generalized version shown in Fig.~\ref{sdtr} below.

@@

Ratings are accommodated by introducing additional criteria and by
modelling a combined response $y$, which represents both the binary
classification decision and rating as a single number. The value of
$y$ increases with the strength of evidence in favor of the second
stimulus class. For example, if confidence is rated on a four-point
scale, then $y = 1$ when a subject decides that a stimulus belongs to
the first class with certainty $4$, and $y = 5$ when a subject decides
that a stimulus belongs to the second class with certainty $1$. More
formally, if $K$ is the number of possible combined responses, then a
subject is assumed to give response $k$ if $s \in
(\bm{c}_{k-1},\bm{c}_k]$, where $\bm{c}$ are the decision criteria,
  with $\bm{c}_0$ and $\bm{c}_K$ fixed at $-\infty$ and $+\infty$
  respectively.

There is a good reason to collect the ratings and use the generalized
SDT model from Fig.~\ref{sdtr}, even when neither the ratings nor the
placement of criteria are relevant to the research problem. When $K =
2$ (no ratings), the SDT model fits perfectly, regardless of whether
it is a reasonably good approximation to reality, because the data and
the model have the same dimensionality. This makes the generalization
to the $K > 2$ case (i.e., decision with ratings) particularly
important, as it is only when $K > 2$ that the formal assumptions of
the model (e.g., equal or unequal variance) can be tested\footnote{In
  contrast to the formal assumptions, a psychological interpretation
  of the SDT parameters can be tested even when ratings are not
  available, e.g., by means of selective influence
  \cite{sternberg2001separate}}, which is often done by comparing
observed and predicted ROC curves.

@@

If data have a hierarchical structure, but variability due to
subjects, items, or other sampled factors is not accounted for,
estimates of average (fixed) effects are not guaranteed to be unbiased
and conclusions are not guaranteed to generalize to the target
population. The not uncommon practice of analyzing data aggregated
over sampled factors represents an extreme case of ignoring
hierarchical data structure. The invalidity of this approach in the
context of SDT was clearly illustrated with the results of
simulational studies by \citeA{morey2008problematic}; however,
strictly speaking, such demonstrations are irrelevant to prooving the
invalidity in question. Because SDT is a non-linear model, by
definition, when estimates of its parameters are based on data
aggregated over sampled factors (e.g., $d'$ estimated for hits and
false alarms averaged or summed over subjects), the expected values of
these estimates (e.g., what the calculated $d'$ actually estimates)
are not in general true population averages (e.g., true $d'$ in some
condition). Such estimates are biased and inference about a target
population based on them is not valid. To give a concrete example,
consider two unbiased participants, one with $d'_1 = 2$ and one with
$d'_2 = 4$. Their expected average accuracy is given by $(\Phi(d'_1/2)
+ \Phi(d'_2/2)) / 2 = .91$, which corresponds to $d' = 2.68$, whereas
their true average $d'$ is $3$.

@@

Firstly, the standard errors or credible intervals associated with the
regression coefficients do not reflect the uncertainty in the SDT
parameter estimates because the latter are treated as mere data
points. The precision of parameter estimates often varies between
participants, items, or conditions, but when the estimates are treated
as data points, no use is made of this information. Secondly,
regressing parameters on numeric predictors makes their estimates
dependent on the common regression structure, and so also on each
other, which can improve the quality of the estimates, just as
assuming that random effects are samples from a common distribution
may improve their estimates. The aforementioned problems can be dealt
with by supplementing the SDT model with the hierarchical linear
regression structure.

\section{Comparison with other software implementations}
% \section{Hierarchical Signal Detection Theory in a constrained
%   parameter space}

Both $d'$ and $c$ have the virtue of being directly interpretable in
terms of sensitivity and bias. However, both $d'$ and $c$ are
constrained: $d'$ is non-negative and, when there is more than one
criterion, the elements of the $\bm{c}$ vector are order restricted
($\bm{c}_{i+1} > \bm{c}_i$). Hierarchical linear regression structure
can only be defined on unconstrained parameters, because random
effects are assumed to be normally distributed and normal distribution
is unbounded and because fixed effects are represented by
unconstrained parameters. We provide examples of the problems that may
arise when the SDT model is not correctly reparametrized in the
following summary of two hierarchical SDT implementations. One is the
Gibbs sampler proposed by \citeA{morey2008problematic} and the other
is the Hierarchical Meta-d' model (HMeta-d') proposed by
\citeA{hmetad}. After describing the problems associated with those
two software implementations we will briefly explain how our method
compares to what the otherwise excellent \code{brms} package has to
offer.

The HMeta-d' model is a hierarchical version of the
meta-d' model \cite{maniscalco2012signal}, which in turn is a
generalization of the SDT model that allows for a separate
"meta-sensitivity" to account for possible discrepancies between a
binary stimulus classification (referred to as a type 1 task) and the
associated rating task (referred to as a type 2 or meta-cognitive
task). We consider HMeta-d' here because it reduces to the SDT model
with ratings when the type 1 and type 2 sensitivities are equal.

The Gibbs sampler created by \citeA{morey2008problematic} allows for
at most two sampled factors to have independent normally distributed
random effects on the evidence distribution means. Unlike $d'$, each
evidence distribution mean considered in isolation is an unconstrained
parameter, but the mean of the second evidence distribution is by
definition greater than ($d' > 0$) or equal to ($d' = 0$) the mean of
the first. The authors explicitly admit that their algorithm does not
enforce this restriction because, as they say in the paper, it would
greatly complicate analysis, although nothing more is said about these
complications. At the same time the authors fail to mention that an
immediate consequence of allowing for negative $d'$ values is that the
resulting posterior distribution loses its intended interpretation
since it can have a non-zero mass when $d' < 0$. The outermost
criteria are fixed at $0$ and $1$, and the ordering restriction is
enforced by assuming that the likelihood is $0$ whenever $\bm{c}_{i+1}
\leq \bm{c}_i$. As the authors explain, because a sampled factor can
have independent random effects on the evidence distribution means, it
can have an effect on all the criteria: shifting both means by the
same amount in the same direction is equivalent to keeping the
sensitivity intact, while shifting the criteria relative to the
evidence distributions. However, the elements of the criteria vector
cannot be affected differently by the same sampled factor, which is an
unrealistic restriction: participants differ in how they place the
criteria just as they differ in their sensitivity.

In HMeta-d' the hierarchical structure is restricted to normally
distributed random intercepts of one sampled factor. In the HMeta-d'
model the $d'$ parameter is allowed to assume negative values also,
but the most problematic aspect of this implementation is again the
representation of the criteria. Each element of the criteria vector
has an associated independent normal distribution, which does allow
for criteria random effects, but does not enforce the necessary
ordering restriction. The elements of this vector are sorted to obtain
another criteria vector, and it is this sorted criteria vector that is
used to model the conditional combined response
distributions. Consequently, the model does contain parameters
representing the actual, order-restricted criteria, but, because
sorting is not injective, the space of the actual criteria is only
loosely related (i.e., not isomorphic) to the space of the
unrestricted criteria vectors that are associated with the
hierarchical structure. This makes posterior distributions of random
criteria effects uninterpretable.

% PORÓWNANIE Z brms
Some extensions of the SDT model can be fitted correctly using the
excellent brms package, as described in \citeA{burkner2017brms}. The
brms package is a flexible tool that shares some deep design
similarities with our method. Both our package and the brms belong to
a growing family of software tools that aim to provide a somewhat
simplified and domain or application specific interface to one of the
general purpose bayesian inference engines, in this case the stan
modelling language \cite{carpenter2016stan}. The \code{brms} is a
highly flexible, well documented package that offers an elegant
interface for fitting a large class of mixed models. Among the models
that can be fitted using this package are the ordinal regression
models. The hierarchical SDT model with ratings is essentially a
generalization of mixed ordinal regression, since ratings are an
ordinal-scale variable. However, that does not mean that a general
hierarchical SDT model with ratings can be fitted using the
\code{brms} package. There are three categorical distributions
available at present in brms, i.e., the cumulative model, the adjacent
category model and the sequential model, each with it's own set of
link functions. As the author of the package explains
\cite{burkner2017brms}, the only model that respects the ordering of
the thresholds is the cumulative model. The thresholds in the
cumulative model cannot be regressed on additional variables,
consequently they cannot be associated with random effects.

Another important difference between our method and the \code{brms}
package is that \code{brms} does not utilize the fact that for ordinal
regression models the data can often be significantly aggregated
without loosing any information. For example, if a basic SDT model
with ratings was to be fitted to an arbitrary large dataset using our
method the aggregated dataset would consist of a matrix of dimension
$2\times K$. In that case each iteration of the sampling algorithm
implemented in stan will consist of essentially 2 computations of the
likelihood function, whereas without any aggregation step the number
of such computations will be a linear function of the sample
size. This is especially problematic for high-dimensional non-linear
models such as hierarchical SDT with ratings, because they require
relatively large datasets (i.e., many data points per participant
$\times$ condition).

@@

Note that the normal distribution centered at the midpoint is merely a
mapping device, not a third evidence distribution, and that, for the
reasons that will be explained later, it is wider than the two
evidence distributions. The mapping expressed by Eq.~\ref{eq:1} is an
isomorphism between the $R^{K-1}$ space and the space of
order-restricted criteria vectors. It's inverse is given by
$\gamma_i =\ln{(\int_{c_{i-1}}^{c_i} f(s) \,ds /
  \int_{c_{K-1}}^{\infty} f(s) \, ds)}$, where $f$ is the standard
normal probability density function. The elements of the $\bm{\gamma}$
vector correspond to relative distances between pairs of adjacent
criteria because their exponents represent the relative magnitudes of
areas under the standard normal curve, delineated by the pairs of
adjacent criteria:
$e^{\bm{\gamma}_i} / e^{\bm{\gamma}_j} = (\Phi(\bm{c}_i) -
\Phi(\bm{c}_{i-1})) / (\Phi(\bm{c}_j) - \Phi(\bm{c}_{j-1}))$. When
$K=2$, only $\bm{\gamma}_1$ is free to vary, and its value directly
represents the direction and magnitude of bias: $\bm{\gamma}_1$ is $0$
when the criterion is placed at the midpoint between the evidence
distributions; the more negative (positive) $\bm{\gamma}_1$ is, the
more the criterion is shifted to the left (right) of the midpoint.

@@

% BOOTSTRAPOWANIE PRIORÓW NA PODST VARIACYJNEGO BAYESA

A Bayesian model is not complete without providing fixed values of all
the parameters that define prior distributions. Specifying the priors
on sensitivity effects does not pose any special difficulties. The
sensitivity of an unbiased classifier given percent correct (pc) is
given by $2 \Phi^{-1}(\text{pc})$. When $p(stim = 1) = 0.5$, the
greater the bias, the lower the accuracy, meaning that an unbiased
sensitivity is a lower bound on sensitivity given percent correct. Let
us assume that the majority of subjects are expected to achieve
percent correct within the $.51$ to $.99$ range, with negligible
bias. Since $\ln(2\Phi^{-1}(.51)) = -2.99$ and
$\ln(2 \Phi^{-1}(.99)) = 1.54$, a reasonable weakly informative prior
on $\delta$ is normal with mean $(1.54 - 2.99) / 2$ and standard
deviation $(1.54 + 2.99) / 2$, which is the default prior on delta
effects in the \code{bhsdtr} package.

@@

The \code{bhsdtr} package implements the model in the Stan modelling
language because the stan sampler uses a state-of-the-art adaptive
Hamiltonian Monte Carlo algorithm which often handles high-dimensional
correlated posteriors better than a Gibbs sampler.

Our package is essentially a collection of documented functions: The
\code{aggregate\_responses} function aggregates data as much as
possible for efficiency, but without distorting the hierarchical
structure. The \code{make\_stan\_model} function creates a model
definition in the Stan language. The Stan code produced by the
\code{make\_stan\_model} function can be fitted as is or modified by
the user if needed, e.g., to change the prior distributions or to drop
the equal variance assumption. The \code{make\_stan\_data} function
creates regression model matrices and other data structures required
by the model created using the \code{make\_stan\_model}
function. Finally, the \code{plot\_sdt\_fit} function can be used to
visually assess the fit of the model by creating publication-ready ROC
curve plots or response distribution plots with posterior predictive
intervals calculated for the chosen $\alpha$ level.

@@

The essential steps of a typical data analysis process will usually
involve preparing the data, creating the model code, fitting the
model, assessing the fit, and possibly converting the unconstrained
$\delta$ and $\gamma$ parameters to $d'$ and $c$.

@@

A model is fitted using the \code{stan} function from the \code{rstan}
package. This function requires a special list of data structures used
by the model as well as a model specification expressed in the Stan
language.

Every model has some fixed effects structure since, even when there
are no predictors, the model parameters can be expressed as regressed
on a vector of ones (i.e., an intercept). However, many models also
have a hierarchical structure and, if that is the case, this
hierarchical structure has to be specified when using the
\code{make\_stan\_model} function. This is done by providing a list of
lists of R model formulae. Each list of model formulae is composed of
at least three elements and specifies the correlated random effects of
one sampled factor. The \code{group} element specifies the sampled
factor; the \code{delta} and \code{gamma} elements specify which
effects are assumed to vary between the levels of this sampled
factor. When \code{make\_stan\_model} is used without any arguments,
it specifies a model without any random effects. Fixed effects model
matrices are specified by providing a list with at least two model
formulae, named \code{delta} and \code{gamma}, to the
\code{make\_stan\_data} function that is described later in this
paper. Non-default priors can be specified by adding optional elements
to the random and fixed effects specification lists, as described in
the \code{make\_stan\_data} function documentation.

In the study in question there was only one sampled factor, i.e., the
participants. Because duration was a within-participant variable, in
principle its effect could vary between the participants for all the
SDT parameters. However, a preliminary data analysis indicated that
the 32 ms difference in duration seemed to affect only the
sensitivity. Thus, it was assumed that $\delta$ may depend on duration
and order, but $\gamma$ may only be affected by order. Because
duration was a within-subjects variable, its effect on $\delta$ was
assumed to vary between the participants, but the only random effect
associated with $\gamma$ was the intercept.

@@

The \code{make\_stan\_data} function creates fixed and random effect
model matrices based on the respective model formulae using dummy
contrast coding. Note that the implicit intercept was removed for the
$\delta$ model matrix (the $-1$ term on the right hand side of the
model formula). In this way, $\delta$ was estimated for every duration
$\times$ order condition. The resulting separate intercepts and slopes
parametrization makes it easier to calculate arbitrary contrasts on
posterior samples. A more standard parametrization was used for the
$\gamma$ parameter because it was initially assumed that the criteria
depend only on order, which is a two-level factor, and so there was
only one contrast of interest for every element of the $\gamma$
vector. On the other hand, even in such cases the parametrization with
separate intercepts and slopes for every condition may be more
convenient if a researcher is interested in the actual criteria, as we
will later explain when introducing the \code{gamma\_to\_crit}
function. This example also illustrates how the separation of the
$\delta$ and $\gamma$ regression structures makes it possible to test
a broad class of linear models representing the dependence of the SDT
parameters on additional variables.

@@

Note that since more than one sampled factor is allowed, the names of
all the hierarchical parameters are indexed (e.g.,
\code{delta\_sd\_1}, \code{delta\_random\_1},
\code{Corr\_delta\_1}). The name \code{counts\_new} refers to
posterior predictive samples that are required by the
\code{plot\_sdt\_fit} function. Names starting with \code{Corr} refer
to random effects correlation matrices.

@@

As can be seen in the previous code fragment four chains of 8,000
iterations each were run simultaneously; the first half of the
posterior samples, which served as a warm-up period for tuning the
parameters of the sampling algorithm, was discarded. Part of the
resulting Stan output is presented in Table~\ref{fitsum} below.

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "paper.toc"
%%% End:
